Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-01-24 papers/LOPe/coarray-comparison.tex
First cut at programming-model (mostly all there) stage.

Below is the motivational example from the original Numrich and Reic paper.
{NOTE: This should sortof/kindof show how it would work between nodes?  Next
section shows OpenMP example for parallelism at the thread level.}

{\small
\begin{verbatim}

subroutine laplace (nrow,ncol,u)
   integer, intent(in) real, intent(inout) real integer
   ! that refers to the current image
   left = me-1
   if (me == 1) left = ncol
   right = me + 1
   if (me == ncol) right = 1
   call sync_all( (/left,right/) ) ! Wait if left and right
   ! have not already reached here
   new_u(1:nrow)=new_u(1:nrow)+u(1:nrow)[left]+u(1:nrow)[right]
   call sync_all( (/left,right/) )
   u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow)
end subroutine laplace

\end{verbatim}
}

Below is the equivalent implementation in LOPe.

{\small
\begin{verbatim}

pure CONCURRENT subroutine laplace(U)
   real, intent(inout), HALO(-1:*:1,-1:*:1) :: U(0:,0:)

   U(0:0) = U(-1,0) + U(+1,0) + U(0,-1) + U(0,+1) - 3*U(0,0)

end subroutine laplace

\end{verbatim}
}

NOTE: Following not a criticism of CAF as CAF is a general purpose parallel
programming language and LOPe only pertains to the halo parallel pattern.

\subsubsection{Improvements}
\begin{itemize}

@@

\item
LOPe applies the transfer of halo memory across multiple (possibly) levels of
memory with the intrinsic {\tt transfer\_halo} function (not shown here).  Thus
the LOPe algorithm can be run on a machine with many interconnected nodes, each
containing hybrid processor cores.  The coarray example can only be run on
multiple nodes (called images in Fortran) without accelerator cores.

\item
The algorithm implementation is separate from user-specified synchronization,
e.g., {\tt call sync\_all}.  In LOPe, synchronization is subsummed in the
semantics of the CONCURRENT attribute and the {\tt transfer\_halo} function.

\item
The algorithm implementation is separate from any specification as to where the
array memory is located.  The CAF example explicity denotes where memory is
located with the {\tt [left]} and {\tt [right]} syntax where left and right
specifiy a processor topology.

\item
The algorithm implementation is separate from any specification as to where the
algorithm is to be executed.  The CAF example explicity denotes where a statement is to
executed with control flow construct like {\tt if (me == 1)}.

@@

\item
The semantics of LOPe makes explicit management of array temporaries (e.g., {\tt
  u} and {\tt new\_u} by programmers unnecessary (though still possible).
Because in LOPe the halo region is a language construct, the compiler is better
able to manage temporary buffers than users on the target hardware platform.

@@

\subsubsection{Errors that are constrained by the language}
\begin{itemize}

@@

\item
A programmer can't make indexing errors in a concurrent routine by going out of
bounds of the array plus halo memory.  The compiler is able to catch this class of
error at compile time unless the halo shape is assumed (e.g., HALO(:,:)).

@@

\item
A programmer can't make synchronization errors as synchronization is implicit in
the CONCURRENT attribute.  A thread running a CONCURRENT procedure is provided
with a copy of it's array element (plus halo) that is consistent with the state
of memory at the time of invocation of procedure.  Stores to an individual
thread's array element (by that thread) is never visible to other threads.
It is possible to relax this restraint with a halo synchronization intrisic.
However, LOPE encourages the creation of small functions and lets the compiler
fuse the procedures together and provide the necessary synchronization.

@@

Please note that this example is somewhat unfair because in practice CAF is usually
refactored in a locally oriented way by so that communication and
synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.

This is the normal way that large complex MPI and CAF programs are implemented, i.e., using
halo cells.  LOPe proposes to formalize this common pattern into the Fortran
language providing the compiler information in order to spread computation over
more hardware resources and reduce complexity for the programmer.

>>>>

LOPe provides a purely \emph{local} viewpoint; the programmer is only provided read and write access
to the local array element and write access to a small halo region surrounding the local element.
There is simply \emph{no} mechanism provided for the programmer to even know \emph{where} the local
element is in the context of the broader array.  On a distributed memory architecture, the halo
elements may not even be physically located on the same processor.  If executed on a cluster
containing hybrid processing elements (e.g. GPUs), the halo elements may be as far as three hops
away: one to get to the host processor and another two to get to memory on the hybrid processor
executing on another distributed memory node.  LOPe provides a complete separation between algorithm
development and memory management and synchronization (between memory copies of the same logical
array region covered by halos).  By explicitly describing the existance and size of an array's halo
region, the compiler is provided with enough information to manage most of the hard and detailed work
involved in memory transfer and synchronization.

Additionally, the semantics of the LOPe execution model (on which more will be said later)
remove even the \emph{possibility} of race conditions developing during execution of a concurrent
procedure.

We emphasize some of these advantages by compairing the \texttt{Laplace} implementation shown above
with the implementation of the same algorithm from the original Numrich and Reid paper describing
coarrays in Fortran.  We should point out that this comparison is somewhat unfair because
Numrich and Reid were describing the advantages of coarray notation for transferring memory
on distributed memory architectures, not how ideally to use coarrays within a major code.  But
this example serves to highlight the advantages of LOPe as described.  In the coarray example
shown below type declarations have been removed to save space:
{\small \begin{verbatim}
subroutine Laplace (nrow,ncol,U)
   left = me-1     ! me refers to the current image
   if (me == 1) left = ncol
   right = me + 1
   if (me == ncol) right = 1
   call sync_all( [left,right] ) ! Wait if left and right have
                                 ! not already reached here
   new_u(1:nrow)=new_u(1:nrow)+u(1:nrow)[left]+u(1:nrow)[right]
   call sync_all( [left,right] )
   u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow)
end subroutine
\end{verbatim}}

The advantages of LOPe over this example are now described.  Please note the the following
is not a criticism of Coarray Fortran (CAF) as CAF is a general purpose parallel programming
language and LOPe only pertains to the halo pattern useful in stencil-based algorithms.

%%Please note that this example is somewhat unfair because in practice CAF is usually refactored in a locally oriented way by so that communication and synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.
\subsubsection{LOPe advantages.}
\begin{itemize}

@@

\item
LOPe applies the transfer of halo memory across multiple (possibly) levels of memory with
the LOPe intrinsic \texttt{TRANSFER\_HALO} function (described later).  Thus the LOPe
algorithm can be run on a machine with many interconnected nodes, each containing hybrid
processor cores.  The coarray example can only be run on multiple nodes (called images in
Fortran) without accelerator cores.

\item
The algorithm implementation is separate from user-specified synchronization, e.g.,
\texttt{call sync\_all}.  In LOPe, synchronization is subsummed in the semantics of the
\texttt{CONCURRENT} attribute and the \texttt{TRANSFER\_HALO} function described later.

\item
The algorithm implementation is separate from any specification as to where the array
memory is located.  The CAF example explicity denotes where memory is located with the
\texttt{[left]} and \texttt{[right]} syntax where left and right specifiy a processor
topology.

\item
The algorithm implementation is separate from any specification as to where the algorithm
is to be executed.  The CAF example explicity denotes where a statement is to executed
with control flow construct like \texttt{if (me == 1)}.

@@

\item
The semantics of LOPe makes explicit management of array temporaries (e.g., \texttt{U} and
\texttt{new\_U} by programmers unnecessary (though still possible).  Because in LOPe the
halo region is a language construct, the compiler is better able to manage temporary
buffers than users on the target hardware platform.

@@

\subsubsection{Errors that are constrained by the language.}
\begin{itemize}

@@

\item
A programmer can't make indexing errors in a concurrent routine by going out of bounds of
the array plus halo memory.  The compiler is able to catch this class of error at compile
time as long as compile-time constants are used to specify halo sizes.

@@

\item
A programmer can't make synchronization errors as synchronization is implicit in
the \texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided
with a copy of it's array element (plus halo) that is consistent with the state
of memory at the time of invocation of procedure.  Stores to an individual
thread's local array element (by that thread) is never visible to other threads.
LOPE encourages the creation of small functions and lets the compiler
fuse the procedures together to provide the necessary synchronization.

@@

### 2015-01-24 papers/LOPe/programming-model.tex
First cut at programming-model (mostly all there) stage.

The second keyword employed by LOPe (although in this case extended) is \texttt{CONCURRENT}.  The
\texttt{concurrent} keyword already exists in the form of a \texttt{do concurrent} loop construct (a
Fortran 2008 feature).  By using \texttt{do concurrent}, the programmer asserts that specific
iterations of the loop body may be executed by the compiler in \emph{any order;} even
\emph{concurrently.}  For example, the following,
\begin{verbatim}
  pure CONCURRENT subroutine Laplace(U)
     real, HALO(:,:) :: U
     U(0,0) =                 U(0,+1)               &
              +  U(-1,0)  - 3*U(0, 0)  +  U(+1,0)   &
                          +   U(0,-1)
  end subroutine Laplace
\end{verbatim}
defines a function that can be used as part of an iterative solution of Laplace's equation in two
dimensions.  A function with the \texttt{pure} and \texttt{CONCURRENT} attributes may be called from
with a \texttt{do concurrent} loop.  One should imagine that this function is called \emph{for each}
\texttt{(i,j)} iterate of the loop.

\subsubsection{LOPe index notation.}
In the \texttt{Laplace} example shown above, the \texttt{(0,0)} array element is the \emph{local} 
array element and only the local element may be modified.

@@

\subsection{Locally-Oriented Programming}

%%We highlight the Fortran elemental source-code abstraction because it provides a local orientation similar to that of OpenCL or CUDA kernel functions that were designed to exploit streaming, highly-threaded architectures like GPUs.  Elemental functions are pure in that they are guaranteed to be free of side effects such as I/O.  Consider a convolution where a 3x3 filter is applied to individual pixel elements in a photograph to average of blur the original photograph.  A convolution is similar to stencil operations in computational fluid dynamics that are used to obtain spatial derivatives of state variables.  Simple code is shown with extentions to Fortran shown in capital letters.  The extensions could be provided to C or Fortran functions via compiler directives rather than explicit language syntax.

Consider a simple average over a five point stencil in two dimensions.  The algorithm updates the
stencil interior point $A(0,0)$ as shown in the following:

The

\begin{verbatim}
   real, allocatable, HALO(-1:*:1) :: A(:)[:], B(:)[:]

   !! Allocate local memory
   !  -------------------------------------------
   allocate(A(N)[*], B(N)[*])
\end{verbatim}

\begin{verbatim}
   device = get_subimage(1)          ! THIS_IMAGE() if device unavailable

   !! Allocate memory on subimage if available
   !  ----------------------------------------
   if (device /= THIS_IMAGE()) then
      allocate(U1h(0:J)  [device], HALO=(-1:*:1))   [[device]]
      allocate(U2h(0:J/2)[device], HALO=(-1:*:1))   [[device]]
   end if

   !! Initialize and transfer memory
   !  ------------------------------
   do j1h = 0, J
      U1h(j1h) = j1h
   end do
   U1h[device] = U1h

   do concurrent (j2h = 1:J/2-1)  [[device]]
      call restrict(U1h(2*j2h)[device], U2h(j2h)[device])
   end do

   do concurrent (j2h = 1:J/2-1)  [[device]]
      call restrict(U1h(2*j2h)[device], U2h(j2h)[device], [0.25, 0.5, 0.25])
   end do

   deallocate(U1h, U2h)

   if (device /= THIS_IMAGE()) then
      deallocate(U1h[device], U2h[device])   [[device]]
   end if

end program
\end{verbatim}

%%where {\tt A} and {\tt Avg} are arrays, but where the {\tt a} argument in the {\tt convolve} definition is declared as a scalar.  The {\tt HALO} function returns a copy of the 3x3 region of {\tt a} and its surrounding neighbors.  The {\tt convolve} function is free of race conditions because of the copy semantics of {\tt HALO} and the implied synchronization of the {\tt CONCURRENT} attribute, whereby no output variables can be updated before all threads have completed execution.  In addition, while any thread may load from an extended region about \emph{its} element with the {\tt HALO} function, it may only store into its own elemental location.

\subsection{Memory Management}

This section describes the hierarchical memory layout and how memory consistency between the 3-levels of memory is maintained.  It describes what is the compiler's responsibility and what is the programmer's responsibility.

%% A note from a conversation with Matt regarding memory management
%%

%The Fortran language has much tighter restrictions on aliasing than does C.
%So unless a variable has the pointer or target attribute, it cannot be aliased.
%Thus the compiler is able to agressively optimized for memory movement between
%the CPU and accelerator.  However, because the design philosophy of coarrays
%is that memory transfer between images can be expensive, the programmer must
%explicitly transfer memory between images with explicit syntax with square
%bracket notation, i.e., $a[1] = a[2]$.  So we allow the compiler to manage
%memory within an image but require the use of {\tt halo\_exchange} for the
%transfer of halo memory between images.

%Fortran currently supports:

%1. Array syntax: e.g., C = A + B, where A, B, C are arrays.  Note that this is implicitly a loop structure, but that no loop indices need be provided.  Also the programmer need not specify where in memory these arrays reside.  Thus this high level syntax allows the compiler more freedom in both memory placement (even across distributed memory nodes) and in runtime code execution (individual array element may be computed by different hardward threads).  This is a simple example and it is a research question as to what compiler directives would be useful for memory placement and other directions to the compiler for efficient code generation.

%2. Pure procedures:  Fortran has syntax for specifying procedures that have no side effects during execution.  Specifying code that is side-effect free code is important information to provide to the compiler so that it can generate efficient multi-threaded code.

%3. Pure elemental procedures: Fortran has syntax for specifying procedures that take only scalar arguments, but may be applied across array elements.  Elemental procedures are ideal for writing code to be executed within a hardware thread.  They resemble OpenCL kernels, but are simpler because they leave all indexing up to the compiler.

%We have determined that additional syntax is needed, in addition to the three language features described above, to allow programmers the ability to express code in Fortran to be targeted for multi-threaded hardware architectures like GPUs.  This additional syntax is provided by functions that return a copy of a small region of memory surrounding an array element (as seen within an elemental procedure) and with functions for thread synchronization.  This additional syntax will allow pure procedures to perform stencil and other convolution-like operations on a copy of memory, synchronize, then store the computed results back to the array element associated with the given thread.

%In addition Fortran has syntax like the target attribute the specifies when variables can be aliased.  This allows for much easier program analysis as the compiler knows that ordinary variables cannot be aliased.  Fortran also has excellent facilities for interoperability with C so that programming in a mixed language invironment is easily accomplished, including interoperability with native Fortran arrays.

>>>>

The second keyword employed by LOPe is \texttt{CONCURRENT}.  The \texttt{concurrent} keyword already
exists in the form of a \texttt{do concurrent} loop construct (a Fortran 2008 feature).  By using
\texttt{do concurrent}, the programmer asserts that specific iterations of the loop body may be
executed by the compiler in \emph{any order;} even \emph{concurrently.}  For example, the following,
\begin{verbatim}
  pure CONCURRENT subroutine Laplace(U)
     real, HALO(:,:) :: U
     U(0,0) =                 U(0,+1)               &
              +  U(-1,0)  - 3*U(0, 0)  +  U(+1,0)   &
                          +   U(0,-1)
  end subroutine Laplace
\end{verbatim}
defines a function that can be used as part of an iterative solution of Laplace's equation in two
dimensions.  A function with the \texttt{pure} and \texttt{CONCURRENT} attributes may be called from
with a \texttt{do concurrent} loop.  One should imagine that this function is called \emph{for each}
\texttt{(i,j)} iterate of the loop.  An example of this usage will be given later in the text.

\subsubsection{LOPe index notation.}
In the \texttt{Laplace} example shown above, the \texttt{U(0,0)} array element is the \emph{local}
array element and only the local element may be modified.  The default zero-based indexing scheme
for the local array element is somewhat different that normal Fortran notation in that by default,
array indices start a 1.  For LOPe, this default convention is modified; indices for arrays declared
with the \texttt{HALO} attribute (within the context of a \texttt{CONCURRENT} procedure) range (in
one dimension) from \texttt{-M} to \texttt{+N} where the halo attribute has been declared as
\texttt{M:*:N} by the originating type declaration.  The other array elements (in the halo region of
the example) are \texttt{U(-1,0)} and \texttt{U(+1,0)} (left and right of local) and
\texttt{U(0,-1)} and \texttt{U(0,+1)} (bottom and top of local).  The geometric positioning of the
array elements can be seen by the arrangement of the expressions on the right-hand side of the
\texttt{Laplace} example.

@@

Note that the use of halo cells is the normal way that large and complex MPI and CAF
programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran
language thus providing the compiler information in order to spread computation over more
hardware resources and reduce complexity for the programmer.
